A Tale of Two Dictionary Learning Problems

### 2020-08-07 manuscripts/thesis/discussion.tex
l1

\subsection{Dictionary learning via $\ell_1$-norm minimization}

A commonly applied workaround to the intractability of Prob.~\ref{SLCopt} (see \cite{tillmann2015computational}) is to swap out the $\ell_0$-norm in \eqref{minsum} for $\ell_1$-norm, thereby transforming the inference of sparse $\mathbf{\overline x}_i$ for a fixed $\mathbf{B}$ into a convex optimization solvable by a linear program. A major advance in compressive sensing was the discovery that \eqref{minsum} can in fact be solved in this way for fixed $\mathbf{B}$ provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}.

The current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ (of bounded column-norm) that solve Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - C_1\varepsilon \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies \eqref{SparkCondition}. Thus, there is some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to the convexified problem as well. It is an open question as to whether there exist practical constraints that would exclude any remaining solutions to the convefixied problem that don't also solve Prob.~\ref{OptimizationProblem}.

>>>>

\subsection{Dictionary learning via convex optimization}

A commonly applied workaround to the intractability of Prob.~\ref{SLCopt} (see \cite{tillmann2015computational}) is to relax the sparsity measuring norm $\|\mathbf{x}\|_0$ to be $\|\mathbf{x}\|_1$, thereby transforming the inference of sparse $\mathbf{\overline x}_i$ for a fixed $\mathbf{B}$ into a convex optimization solvable by a linear program. Recalling the definition of $\mathbf{z}_i$ in \eqref{LinearModel}, the approach is rather:

\begin{problem}\label{ConvexifiedOptimizationProblem}
Find a matrix $\mathbf{B}$ and vectors \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} solving:
\begin{equation}\label{l1minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_1 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{equation}
\end{problem}

A major advance in compressive sensing was the discovery that, when $\mathbf{B}$ is held fixed, solutions to \eqref{l1minsum} are in fact solutions to \eqref{minsum} provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}. Interestingly, the current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ (of bounded column-norm) that solve Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - C_1\varepsilon \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies \eqref{SparkCondition}. Thus, there is some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to the convexified problem as well. It is an open question as to whether practical constraints would exclude any remaining solutions to the convefixied problem that don't also solve Prob.~\ref{OptimizationProblem}.

As an intuitive example to kickstart these efforts, consider the simple noiseless ($\eta = 0$) case for $k=1$ where the matrix $\mathbf{A}$ has linearly independent unit-norm columns and $\mathbf{x}_i = c\mathbf{e}_i$ for some $c \geq 1$ for all $i \in [m]$. Suppose that for some $\mathbf{B}$ also having linearly independent unit-norm columns there are vectors $\mathbf{\overline x}_i$ satisfying $\|\mathbf{\overline x}_i\|_1 \leq c$ and $\mathbf{B \overline x}_i = \mathbf{Ax}_i$ for all $i \in [m]$. Intuitively, the only vectors in the convex polygon $\{ \mathbf{B \overline x}: \|\mathbf{\overline x}\|_1 \leq c\}$ for which $\|\mathbf{B \overline x}_i\|_2 = \|\mathbf{Ax}_i\|_2 = \|c\mathbf{Ae}_i\|_2 = c$ are its corners $\mathbf{B}(\pm c\mathbf{e}_j)$ for $j \in [m]$, so the $\mathbf{\overline x}_i$ must also be 1-sparse and we can apply Thm.~\ref{DeterministicUniquenessTheorem}.

%Alternatively, it may be the case that $L_{2k}(\mathbf{B}) > 0$ is a necessary constraint to impose to guarantee recovery of $\mathbf{A}$ by solving \eqref{l1minsum}.

### 2020-08-07 manuscripts/thesis/introduction.tex
l1

It is a longstanding practice in the field of signal processing to describe signals as linear combinations of elementary waveforms selected from a pre-specified ``dictionary". Every signal has a unique representation in terms of these atomic components when the dictionary forms what is called a basis. For example, a signal can be decomposed into its constituent frequencies via the Fourier transform, a linear change of basis. %When the basis is orthonormal, the representational coefficient scaling a given elementary waveform is simply the inner product of that waveform with the signal. Equally simple decomposition rules apply when the basis is non-orthogonal. 
%When non-orthogonal, the coefficients are computed by taking the inner product with respect to the components of the dictionary inverse.
Bases have been the dominant form of signal representation until recently, due largely to their simplicity.

For many signal analysis tasks, however, no one basis can convey clearly all of the relevant information in the signal. For instance, if a signal can be either a sine wave or a delta function, then neither the standard basis nor the Fourier basis will indicate one case as explicitly as it does the other. 
The need for such freedom of expression led to the development of redundant signal representations utilizing overcomplete dictionaries, which consist of more atoms than there are dimensions of the signal. An overcomplete dictionary admits infinitely many possible ways to decompose a signal into a linear superposition its constituent atoms, and the intention is to seek the optimal such representation as measured by some task-specific cost function.

A popular approach to the design of overcomplete dictionaries has been to seek one which admits a sparse representation for every signal of interest; that is, each signal can be represented, or at least well-approximated, as a combination of only a few dictionary elements from the bunch. Continuing with our example, the union of the standard basis with the Fourier basis is an overcomplete dictionary with respect to which both sine waves and delta functions (or any finite superpositions thereof) achieve the sparsest possible representation.

Early approaches to sparse representation modeling assumed a model of the signal class from which a suitable sparsifying dictionary could be derived, as we have in our trivial example. Such dictionaries are typically characterized by an analytic formulation and a fast implicit implementation; yet they tend to be over-simplistic models when applied to natural phenomena.

@@

%and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, 
In light of these observations, it is natural to wonder to what extent the optimal dictionary can be identified given a representative sample from a signal class. Answers to this question have implications in practice; wherever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}), it is assumed this structure captures some identifiable physical or causal variable.

@@

%injective on a sparse domain
In this work, it is proven very generally that uniqueness and stability is a typical property of learned dictionaries. 
Specifically, if each of $N$ observed $n$-dimensional real signals is truly a (noisy) linear combination of at most $k$ elementary waveforms drawn from a suitable dictionary of size $m \ll N$, that dictionary is uniquely specified up to an error that is linear in the noise given \mbox{$N = m(k-1){m \choose k} + m$} such signals (Thm.~\ref{DeterministicUniquenessTheorem} and Cor.~\ref{DeterministicUniquenessCorollary}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the problem is well-posed, as per Hadamard \cite{Hadamard1902}, given a sufficient amount of data (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). 
Similar guarantees also hold for the related (and perhaps more commonly posed, e.g. \cite{rehnsommer2007}) optimization problem seeking a dictionary minimizing the average number of elementary waveforms required to reconstruct each sample of the dataset (Thm.~\ref{SLCopt}). To great practical benefit (and technical pain!), these guarantees apply without any assumptions at all on learned dictionaries beyond an upper bound on their size, which is necessary in any case to avoid trivial solutions (e.g., allowing $m = N$). %That is, every pair of solutions of comparable size has some number of dictionary elements in common (up to noise), and similarly so for the coefficients of sparse codes they induce.

@@

\begin{problem}\label{OptimizationProblem}
Find matrices $\mathbf{B}$ and vectors \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} solving:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_0 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}

>>>>

It is a longstanding practice in the field of signal processing to describe signals as linear combinations of elementary waveforms selected from a pre-specified ``dictionary". Every signal has a unique representation in terms of these atomic components when the dictionary forms a basis. For example, a signal can be decomposed into its constituent frequencies via the Fourier transform, which performs a linear change of basis. %When the basis is orthonormal, the representational coefficient scaling a given elementary waveform is simply the inner product of that waveform with the signal. Equally simple decomposition rules apply when the basis is non-orthogonal. 
%When non-orthogonal, the coefficients are computed by taking the inner product with respect to the components of the dictionary inverse.
Until recently, bases have been the dominant form of signal representation due largely to their simplicity.

For many signal analysis tasks, no one basis can convey clearly all of the relevant information in the signal. If a signal can be either a sine wave or a delta function, then neither the standard basis nor the Fourier basis will indicate one case as explicitly as it does the other. 
The need for such freedom of expression led to the development of redundant signal representations utilizing overcomplete dictionaries, which consist of more atoms than there are dimensions of the signal. While an overcomplete dictionary admits infinitely many possible ways to decompose a signal into a linear superposition its constituent atoms, the intention is to seek the best such representation, be it by some analytic criteria, or as measured by some task-specific cost function.

A popular approach to the design of overcomplete dictionaries has been to seek one which admits a sparse representation for every signal of interest; that is, each signal can be represented, or at least well-approximated, as a combination of only a few dictionary elements from the bunch. Continuing with our example, the union of the standard basis with the Fourier basis is an overcomplete dictionary with respect to which both sine waves and delta functions (or any finite superpositions thereof) achieve the sparsest possible representation as a scaling of a single elementary waveform from the dictionary.

Early approaches to sparse representation modeling assumed a model of the signal class from which a suitable sparsifying dictionary could be derived, as we have in our trivial example. Such dictionaries are characterized by an analytic formulation and a fast implicit implementation; yet they tend to be over-simplistic models when applied to natural phenomena.

@@

%and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, 
In light of these observations, it is natural to wonder when and to what extent the optimal dictionary for sparse representation modeling can be identified given a representative sample from a signal class. Answers to this question have implications in practice wherever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}), the assumption being that this structure captures some identifiable physical or logically causal variable.

@@

%injective on a sparse domain
In this work, it is proven very generally that uniqueness and stability is a typical property of learned dictionaries. 
Specifically, if each of $N$ observed $n$-dimensional real signals is truly a (noisy) linear combination of at most $k$ elementary waveforms drawn from a suitable dictionary of size $m \ll N$, that dictionary is uniquely specified by the data up to an error that is linear in the noise given \mbox{$N = m(k-1){m \choose k} + m$} such signals (Thm.~\ref{DeterministicUniquenessTheorem} and Cor.~\ref{DeterministicUniquenessCorollary}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the problem is well-posed, as per Hadamard \cite{Hadamard1902}, given enough data (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). 
Similar guarantees also hold for the related (and perhaps more commonly posed, e.g. \cite{rehnsommer2007}) optimization problem seeking a dictionary minimizing the average number of elementary waveforms required to reconstruct each sample of the dataset (Thm.~\ref{SLCopt}). To great practical benefit (and technical pain!), these guarantees apply without the imposition of any constraints at all on learned dictionaries beyond an upper bound on their size, which is necessary in any case to avoid trivial solutions (e.g., allowing $m = N$). %That is, every pair of solutions of comparable size has some number of dictionary elements in common (up to noise), and similarly so for the coefficients of sparse codes they induce.

@@

\begin{problem}\label{OptimizationProblem}
Find a matrix $\mathbf{B}$ and vectors \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} solving:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_0 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}
